Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-07-06 Correspondence/revisions.tex
updated letter
fixed typo
minor comment 6 closes #34
Done responding to comments 2 and 3 of reviewer 1 and described the changes in the letter. Closes #22 Closes #23

\begin{description}
 \item[Reviewer 1:] responses are listed below.

 \begin{enumerate}
  \item A simulation experiment was done, as requested and included as a new 
section in the supplementary material. The simulations shows that the inference 
methodology can recover the exact parameters user to simulate  the data.
\item The text of the methods section was improved to answer the four issues 
raised in this comment.
 \end{enumerate}

 \item [Reviewer 2:]
\end{description}

>>>>

\begin{description}
 \item[Reviewer 1:] Responses to the main comments are listed below. The minor 
comments where also accepted, namely the addition of readme do the code 
repository as well as the citation of two  additional papers.

 \begin{enumerate}
  \item A simulation experiment was done, as requested and included as a new 
section in the supplementary material. The simulations shows that the inference 
methodology can recover the exact parameters user to simulate  the data.
\item The text of the methods section was improved to answer the four issues 
raised in this comment.
\item The introduction and discussion sections were expanded to included the 
additional information required by this comment.
 \end{enumerate}

 \item [Reviewer 2:] \textbf{Minor comments:} Fixed typo in the abstract; 
 \begin{enumerate}
  \item 
 \end{enumerate}

\end{description}

### 2015-07-06 draft_lm1.tex
Done responding to comments 2 and 3 of reviewer 1 and described the changes in the letter. Closes #22 Closes #23

Dengue is an arthropod-borne febrile disease caused by a flavivirus with four 
serotypes which causes an estimated  $50$ million infections each 
year~\citep{Guzman2010}.
In humans, immunity against a particular serotype is considered permanent 
after the exposure and cross immunity to other serotypes is considered short 
lived~\citep{halstead_dengue_2007} although some 
studies argue for a longer duration of cross-immunity \citep{Reich2013, 
salje2012}.
As a consequence, the proportions of viral serotypes co-circulating at any 
point in time are strongly dependent on previous incidence patterns of the 
disease, which determine the number of individuals susceptible to each serotype 
at any point in time.
% LM: Maybe a citation here?

@@

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\citep{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\citep{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
It is interesting to notice that the larger epidemics, in terms of peak size are 
not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$.

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. 
Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic, but the peak of cases
may occur in the following year.
$^\ddag$: percentage of population.
These results show considerable variation in AR between epidemics, consistent with
the accquiring and loss of serotype-specific immunity.}
\begin{center}
\begin{tabular}{l|c|c|c}
\hline
$^\dag$Year & $^\ddag$Cases & median Attack Ratio & $^\ddag$$S_0$ \\
\hline
1996 & 0.066 & 0.39 (0.17-0.54) & 0.171(0.12-0.38)\\
1997 & 0.238 & 0.87 (0.74-0.87) & 0.273(0.27-0.32)\\
1998 & 0.0708 & 0.5 (0.49-0.5) & 0.142(0.14-0.14)\\
1999 & 0.0371 & 0.11 (0.037-0.2) & 0.345(0.18-1.0)\\
2000 & 0.394 & 0.25 (0.24-0.27) & 1.55(1.5-1.6)\\
2001 & 2.38 & 0.48 (0.47-0.49) & 4.95(4.8-5.1)\\
2005 & 0.217 & 0.15 (0.1-0.21) & 1.47(1.0-2.1)\\
2006 & 0.315 & 0.11 (0.08-0.14) & 2.81(2.2-3.7)\\
2007 & 2.03 & 0.15 (0.15-0.15) & 13.5(1.3e+01-1.4e+01)\\
2008 & 0.0923 & 0.14 (0.031-0.31) & 0.672(0.3-2.4)\\
2010 & 0.831 & 0.18 (0.17-0.19) & 4.54(4.3-4.8)\\
2011 & 1.85 & 0.086 (0.082-0.094) & 21.5(2e+01-2.3e+01)\\
2012 & 0.864 & 0.14 (0.13-0.15) & 6.21(5.8-6.8)\\

>>>>

Dengue is an arthropod-borne febrile disease caused by a flavivirus with four 
serotypes (DEN-1, 2, 3 and 4) which causes an estimated  $50$ million 
infections each 
year~\citep{Guzman2010}.
In humans, immunity against a particular serotype is considered permanent 
after the exposure and cross immunity to other serotypes is considered short 
lived~\citep{halstead_dengue_2007} although some 
studies argue for a longer duration of cross-immunity \citep{Reich2013, 
salje2012}.
As a consequence, the proportions of viral serotypes co-circulating at any 
point in time are strongly dependent on previous incidence patterns of the 
disease, which determine the number of individuals susceptible to each serotype 
at any point in time.
% LM: Maybe a citation here?

@@

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\citep{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\citep{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years. Dengue has been 
reintroduced in Rio de Janeiro in 1986 after being absent for 68 
years\citep{nogueira1999dengue}. During the last decades of the 20th century 
only DEN-1 and DEN-2 serotypes

@@

Table~\ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak.
 Underreporting of cases, which is known to exist but of which exact figures 
cannot be determined, will lead to underestimation $S_0$. However, The attack 
ratio shall remain unbiased as the underreporting affects both the numerator and 
the denominator of equation \ref{eq:AR2}.
It is interesting to notice that the larger epidemics, in terms of peak size 
are not the one with the greater attack ratios.
This stresses the importance of knowing the immunological structure of the 
population. Knowing the $S_0$ for the circulating viruses we can order to more 
accurately assess the potential impact of a coming epidemic, since particularly 
virulent types, can be rendered less of a threat by a low $S_0$. Hon\'orio et 
al.\citep{honorio2009spatial} conducted a serological survey in three separate 
localities within the city, right before the 2008 epidemic, the authors report 
seroprevalences varying from 56-77.4\% which is compatible with our prediction 
of 87.5\% ($1-S_{0,2007}$) for the entire city, considering that we 
underestimate $S_0$ due to the underreporting of cases.

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. 
Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic, but the peak of cases
may occur in the following year.
$^\ddag$: percentage of population.
These results show considerable variation in AR between epidemics, consistent with
the accquiring and loss of serotype-specific immunity.}
\begin{center}
\begin{tabular}{l|c|c|c|c}
\hline
$^\dag$Year & $^\ddag$Cases & median Attack Ratio & $^\ddag$$S_0$ & Circulating 
 Serotypes \\
\hline
1996 & 0.066 & 0.39 (0.17-0.54) & 0.171(0.12-0.38)& DEN-1 and 
2\citep{nogueira1999dengue}\\
1997 & 0.238 & 0.87 (0.74-0.87) & 0.273(0.27-0.32)& DEN-1 and 
2\citep{nogueira1999dengue}\\
1998 & 0.0708 & 0.5 (0.49-0.5) & 0.142(0.14-0.14)& DEN-1 and 
2\citep{nogueira1999dengue}\\
1999 & 0.0371 & 0.11 (0.037-0.2) & 0.345(0.18-1.0)&DEN-1, 2 and 
3\citep{de2004dengue}\\
2000 & 0.394 & 0.25 (0.24-0.27) & 1.55(1.5-1.6)&DEN-1, 2 and 
3\citep{de2004dengue}\\
2001 & 2.38 & 0.48 (0.47-0.49) & 4.95(4.8-5.1)&DEN-1, 2 and 
3\citep{de2004dengue}\\
2005 & 0.217 & 0.15 (0.1-0.21) & 1.47(1.0-2.1)&\\
2006 & 0.315 & 0.11 (0.08-0.14) & 
2.81(2.2-3.7)&DEN-2\citep{nogueira2011dengue}\\
2007 & 2.03 & 0.15 (0.15-0.15) & 
13.5(13.0-14.01)&DEN-2\citep{nogueira2011dengue}\\
2008 & 0.0923 & 0.14 (0.031-0.31) & 
0.672(0.3-2.4)&DEN-2\citep{nogueira2011dengue}\\
2010 & 0.831 & 0.18 (0.17-0.19) & 
4.54(4.3-4.8)&DEN-2\citep{nogueira2011dengue}\\
2011 & 1.85 & 0.086 (0.082-0.094) & 
21.5(20.0-23.0)&DEN-1, 2 and 4\citep{nogueira2011dengue}\\
2012 & 0.864 & 0.14 (0.13-0.15) & 6.21(5.8-6.8)&\\
